Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-03-08 index.qmd
starting to revise front end
revising front end v3 continued

::: {.column width="50%"}
**[University of Chicago admissions counselors, IL territories](https://collegeadmissions.uchicago.edu/contact)**
![](./assets/images/chicagoland_territories_u_chicago_redacted.png)
:::
:::

>>>>

::: {.column width="50%"}
**[University of Chicago admissions counselors, IL territories](https://collegeadmissions.uchicago.edu/contact)**
![](./assets/images/territories/chicagoland_territories_u_chicago_redacted.png)
:::
:::

### 2025-03-08 text/front_end_v3.qmd
revising front end v3 continued

\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

@@

@@

@@

@list_empirics say this:

@@

Third, Geomarkets were incorporated into the College Board "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling "names" [@belkin2019-studata; @list_biz]. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. Lists contain contact information for prospective students. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets. An analysis of 830 student lists purchased by 14 public universities found that 16% of purchases filtered on Geomarket [@RN5012].

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than $35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.]

Taken together, the Market Segment Model show a colleges which Geomarkets possess attractrive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS), based on the Market Segment Model [@RN4987; @RN4980]. EPS was an early software-as-service platform that recreated the analyses of @RN4982. For each Geomarket, colleges could obtain Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors. Based on conversations with enrollment management professionals, EPS software also provided information about the score-sending behavior of individual high schools within each Geomarket. Therefore, colleges used EPS software to decide which Geomarkets to recruit from and which high schools to visit within targeted Geomarkets. Typical marketing material describes EPS as,

> The marketing software that pinpoints the schools and Geomarkets where your best prospects are most likely to be found. With the click of a mouse, EPS provides you with comprehensive reports on your markets, your position in those markets, and your competition. Focus your valuable time and resources on the right prospects [@RN5019].

@@

EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."]. Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

START HERE FRIDAY!!!!!

EPS may also weaken the authority of local decision-makers. @RN4982 sought to develop a concrete, data-driving framework -- The Market Segment Model -- replicated the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an enrollment management consultant -- to plan recruiting efforts centrally. In background conversations, enrollment management professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, use of EPS contributed to most colleges visiting the same sets of affluent, high-achieving high schools.[^9]

[^9]: We were told that savvier EPS users visited high schools that EPS recommended not visiting because there would be less competition for the good students that attended these high schools.

Research question 1 asks, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We expect substantial socioeconomic inequality between geomarkets in large metropolitan areas. Given the extent of class- and race-based residential segregation in the U.S., it would be surprising to not observe such inequality.

Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:

We expect substantial racial inequality between geomarkets in large metropolitan areas for several reasons. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race once. U.S. cities are characterized be extreme historic and contemporary residential racial segregation [@RN4779]. Structures built upon racialized structures are racialized structures. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, geomarket borders may have been drawn along class divides. A strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the countours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets. Although these arguments are speculative, the extent of racial inequality between geomarkets can be assessed empirically. We present the following hypothesis:

## Geomarket Filter in Student Search Service

In XXXX, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists. For public universities that purchased lists, 80% purchase more than 50,000 names annually. @RN4896 reports that student lists were the top expenditure item in the undergraduate recruiting budget for both private and public institutions in 2022, with the average public institution allocating 15% of its budget to purchasing names. Historically, dominant list vendors are college board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

Student lists are positively associated with student outcomes. A College Board research report by @RN4739 compared SAT test-takers who opted into the College Board Student Search Service -- allowing colleges to purchase their contact information -- to those who opted out. After controlling for covariates (e.g., SAT score, parental education, school fixed effects), 41.1% of students who participated in Search attended a 4-year college compared to 32.8% of students who opted out, an 8.3 percentage point difference and a 25.3 percent change -- $(41.1–32.8)/32.8$ -- in the relative probability. Participating in Search was associated with a larger change in the relative probability of attending a 4-year college for Black students (24.5%) and Hispanic students (34.4%) than White students (21.6%), and a larger change for students whose parents did not attend college (40.6%) than those whose parents had a bachelor’s degree (18.9%).[^10] @RN4739 also found that participating in Search was strongly associated with obtaining a BA in four years, particularly for sub-groups that were historically excluded from college.[^11]

[^10]: Leveraging a natural experiment in College Board student list purchases, @RN4894 find that purchasing a prospect profile increases the probability that the student will apply to and enroll at the purchasing college, with larger effects for Black, Hispanic, and lowincome students.

@@

However, @RN5012 argue that student list products exacerbate racial inequality in college access. @RN5012 conceptualize student list products as "selection devices" [@RN4778] that enable colleges to select which prospective students they target by incorporating search filters (e.g., high school graduating class, state). @RN4786 defines "racialized inputs" as ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) meet the criteria of racialized inputs. Using a national sample of high school students and using data from actual student lists purchased by public universities, @RN5012 show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. <!-- Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted. -->

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Products built upon structural racism are structurally racist products. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). Additionally, the College Board created geographic borders that subsequently became filters in the *Student Search Service*.[^12] Geomarket search filters, slice states and metropolitan areas into smaller, local recruiting markets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982, nor subsequent promotional material [e.g., @RN1621], mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score. This article extends @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets.

@@

@RN4982 [pp. 42-44] provides recommendations about how The Market Segment Model and Geomarkets should be utilized in the context of purchasing student lists. The Market Segment model "simply demonstrates what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations" (p. 42). The authors describe a college that "draws most of its students from regional or national segments" (p. 42). Focusing on New England -- including the Boston Geomarket, which was relatively low-income in 1980 -- the authors ask, "where would you concentrate your energies" (p. 42),

> Ideally, you would seek communities \[Geomarkets\] with a high proportion of students already predisposed toward institutions such as your own. The Market Segment Model would provide this information through segment percentages for the community in question. Further classification of students by social attributes allows you to identify a group for mailings or recruiting\ldots If you were to recruit in Boston, only about two out of every ten students with fewer than two attributes would likely listen, while slightly less than half of the students with two or more attributes would be receptive\ldots. Your efforts would surely be better directed toward three of the four communities in the bottom band, Manchester, Hartford, and Fairfield County. Simply by knowing a little bit about the students' backgrounds and academic records you could quickly focus your attention on those most likely to consider your kind of institution. Indeed, in Fairfield County alone you could reach more than 40 percent of your "primary target" population -- that is, students with a greater than 75-percent probability of concentrating their college choices among institutions like your own" (p. 42-44)

In both mailings and off-campus recruiting visits, @RN4982 recommend focusing on Geomarkets with large populations of students who are deciding between "institutions like your own" (p. 44). Selective colleges primarily draw from students in the "regional" and "national" segments. @RN4982 suggest that the student list purchases of selective colleges should focus on Geomarkets with large numbers of affluent, highly educated households. By contrast, the student list purchases of local state colleges and community colleges, which rely on enrollment from students in the "local" and "in-state," should focus on nearby Geomarkets with large numbers of low-SES students. Our analyses attempt to recreate the strategy recommended by @RN4982. Considering the correlation between race and wealth in segregated America, we suggest that a class-based list-buying strategy will increase racial stratification in college access. Furthermore, we anticipate that the students of color targeted by this strategy are disproportionately enrolled in affluent, predominantly white schools.

```{=html}
<!-- THIS TEXT IS ABOUT UTILIZATION. CUT FOR NOW.
Finally, we suggest that utilization of geomarket filters within the Student Search Service product is likely to amplify class-based and race-based disparities in college opportunity. In contrast to standardized selection devices (e.g., the FICO score algorithm), student list products are discretionary selection devices in that colleges have discretion of which filters they elect to select prospects. Compared to standardized selection devices, scholars tend to find that administrative discretion over selection devices causes structural inequality to increase [@RN4883; @RN4881; @RN4795]. Even in the absence of explicit or implicit bias, discretionary selection devices can increase racial inequality when decision-makers have incomplete knowledge about how the inputs they choose are correlated with race. For example, research shows that Americans dramatically underestimate the magnitude of racial income inequality [@RN4884]. As such, discretionary selection devices are particularly sensitive to inputs that are correlated with race but are perceived to be race-neutral.

Geomarkets are opaque in that geomarket borders may be unclear to someone purchasing student lists and very few people know how these borders were created. @RN4849 argues that opacity in algorithmic products is intentional and is a core criterion of "weapons of math destruction." Enrollment management professionals utilizing EPS and *Student Search Service* may interpret geomarkets as a race-neutral recruiting tool. When purchasing student lists, @RN4982 [p. X] suggest that affluent colleges should consider targeting affluent geomarkets. Student list purchases do not show how the characteristics of targeted prospects compare to the characteristics of the surrounding community. From this perspective, racial inequality from the use of geomarkets in student list purchases may be the unintentional result of using a seemingly race-neutral product. Furthermore, purchases that specify multiple filters (e.g., test scores, grades, intended major) including geomarkets can yield unintended racial inequality because administrators have incomplete knowledge about how the intersection of these filters interact with local patterns of segregation.
-->
```

```{=html}
<!-- CUT TEXT DEFINING COLORBLIND RACISM AND RACE-NEUTRAL RACISM
Colorblind racism describes ideology and associated behaviors that ignore the significance of race and racial discrimination [@RN5020]. The Market Segment Model can be seen as an example of colorblind racism. @RN4982 develop an analysis of the structure of college choice that ignores race but we contend that Geomarkets are correlated with race. Race-neutral racism is racial inequality that results from using ostensibly race-neutral inputs that are systematically correlated with race [CITE] [e.g., @RN4775; @RN4786; @RN4975]. 
-->
```

```{=html}
<!-- CUT PARAGRAPH: WHY GEOMARKETS ARE WEAPONS OF MATH DESTRUCTION
The incorporation of Geomarkets within the Student Search Service student list product raises concerns about the three characteristics of weapons of math destruction. First, Geomarkets are opaque. The borders of a geomarket may be unclear to someone purchasing student lists and very few people know how these borders were created \[COULD TALK ABOUT PRODUCT UTILIZATION HERE; NOT CLEAR HOW GEOMARKETS INTERACT WITH OTHER STUDENT LIST FILTERS\]. Second, Geomarkets scale. Each year, thousands of colleges purchase student lists and it is easy to filter purchases based on Geomarkets in concert with other filters (e.g., SAT score, GPA). Third, Geomarkets can cause damage. When a college decides to purchase names from one Geomarket but exclude prospects from an adjacent Geomarket, this affects which college students have an opportunity to attend. @RN4982 suggests that selective colleges focus on geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.
-->
```

>>>>

\pagenumbering{roman}
\newpage

@@

\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

@@

On January 9, 2019 a public research university purchased three "student lists" from College Board in order to obtain the contact information of out-of-state prospects, who would subsequently be sent email, mail, and targeted social media. Each purchase utilized three "search filters," thereby controlling which prospects profiles were purchased. First, all three student list purchases filtered for the 2020 high school class. Second, under the heading "College Board Exams," each purchase filtered on a different PSAT score range: 1070 - 1180 for order #448922; 1190 - 1260 for order #448427; and 1270 - 1520 for order #448440. Third, under the heading "Geography", each purchase filtered on the same 67 "Geomarkets," which had titles like "IL07 - Chain of Lakes", "IL08 - Northwest Suburbs", "IL09 - North Shore."

Geomarkets carve states and metroplitan areas into smaller geographic regions meant to define local recruiting markets. A closer look at the Geomarkets selected for orders #448922, #448427, and #448440 shows that each order selected all Geomarkets in the Chicago-land area, except "IL10 - City of Chicago", which is significantly poorer and less White than surrounding Geomarkets. Similarly, in the DC/Maryland/Virginia metro area, each order included the "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but not the "MD05 - Prince Georges Metropolitan" "MD07 - Baltimore (Urban)" or "DC01 - District of Columbia" Geomarkets which are poorer and more Black. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. However, the above Geomarket selections raise the concern that Geomarkets -- and other geographic search filters -- may be means of excluding working class and non-white students from educational opportunities.

This manuscript does blah blah.

@@

@@

Third, Geomarkets were incorporated into the College Board "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling "names" [@belkin2019-studata; @list_biz]. <!-- @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. Lists contain contact information for prospective students.--> The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets. An analysis of 830 student lists purchased by 14 public universities found that X% of purchases filtered on Geomarket [@RN5012].

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than $35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.] Taken together, the Market Segment Profile shows a colleges which Geomarkets possess attractrive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

**Enrollment Planning Service**. In 1984, College Board created the Enrollment Planning Service (EPS), based on the Market Segment Model [@RN4987; @RN4980]. EPS was an early software-as-service platform that recreated the analyses of @RN4982. For each Geomarket, colleges could obtain Market Segment Report for each local market and the Institutional Profile -- their own and that of competitors. Based on conversations with enrollment management professionals, EPS software also provided information about the score-sending behavior of individual high schools within each Geomarket. Therefore, colleges used EPS software to decide which Geomarkets to recruit from and which high schools to visit within targeted Geomarkets. Typical @RN5019 marketing material describes EPS as,

> The marketing software that pinpoints the schools and Geomarkets where your best prospects are most likely to be found. With the click of a mouse, EPS provides you with comprehensive reports on your markets, your position in those markets, and your competition. Focus your valuable time and resources on the right prospects.

@@

EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

EPS may also weaken the authority of local decision-makers. @RN4982 sought to develop a concrete, data-driven framework -- The Market Segment Model -- that replicates the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an enrollment management consultant -- to plan recruiting efforts centrally. In background conversations, enrollment management professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools explicitly not recommended by EPS because there would be less competition for the good students that attended these high schools.

Finally, we suspect that Geomarkets affect how admissions offices organize the recruiting territories assigned to each admissions officer. Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it was common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In those states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago [ADD TO APPENDIX], of Lake Forest College,  of Saint Mary's College [ADD TO APPENDIX], [ADD TO APPENDIX]. <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

Although we cannot gain access to EPS software, Geomarkets are fundamental to EPS software and (arguably) to the organization of recruiting territories for many colleges. Therefore, we are interested in the extent to which this building block is associated with race and class. Research question 1 asks, what is the socioeconomic and racial variation between Geomarkets and how does this variation change over time? We focus on metropolitan areas that are associated with three or more Geomarkets.

We expect substantial socioeconomic inequality between geomarkets in large metropolitan areas. Given the extent of class- and race-based residential segregation in the U.S., it would be surprising to not observe such inequality. Moreover, @RN4982 viewed demand for higher education as a function of class and developed Geomarkets with an eye towards distinguishing geographic areas that differed from one another in terms of class composition and college-going behavior.

<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->

We also expect substantial racial inequality between geomarkets in large metropolitan areas for several reasons. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race once. U.S. cities are characterized be extreme historic and contemporary residential racial segregation [@RN4779]. Structures built upon racialized structures are racialized structures [@RN4786]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, geomarket borders may have been drawn along class divides. A strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the countours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets.

**Geomarket Filter in Student Search Service**. By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service [CITE]. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.^[For public universities that purchased lists, 80% purchase more than 50,000 names annually. @RN4896 reports that student lists were the top expenditure item in the undergraduate recruiting budget for both private and public institutions in 2022, with the average public institution allocating 15% of its budget to purchasing names.] Historically, dominant list vendors are college board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

Student lists are positively associated with student outcomes. A College Board research report by @RN4739 compared SAT test-takers who opted into the College Board Student Search Service -- allowing colleges to purchase their contact information -- to those who opted out. After controlling for covariates (e.g., SAT score, parental education, school fixed effects), 41.1% of students who participated in Search attended a 4-year college compared to 32.8% of students who opted out, an 8.3 percentage point difference and a 25.3 percent change -- $(41.1–32.8)/32.8$ -- in the relative probability.[^10] @RN4739 also found that participating in Search was strongly associated with obtaining a BA in four years, particularly for sub-groups that were historically excluded from college.[^11]

[^10]: Participating in Search was associated with a larger change in the relative probability of attending a 4-year college for Black students (24.5%) and Hispanic students (34.4%) than White students (21.6%), and a larger change for students whose parents did not attend college (40.6%) than those whose parents had a bachelor’s degree (18.9%). Leveraging a natural experiment in College Board student list purchases, @RN4894 find that purchasing a prospect profile increases the probability that the student will apply to and enroll at the purchasing college, with larger effects for Black, Hispanic, and lowincome students.

@@

However, @RN5012 argue that student list products exacerbate racial inequality in college access. @RN5012 conceptualize student list products as "selection devices" [@RN4778] that enable colleges to select which prospective students they target by incorporating search filters (e.g., high school graduating class, state). @RN4786 defines "racialized inputs" as ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) meet the criteria of racialized inputs. Using a national sample of high school students and using data from actual student lists purchased by public universities, @RN5012 show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects.

Because of the extent of residential segregation in the U.S., geographic borders are a commonly studied racialized input in scholarship about algorithmic bias [@RN4775; @RN4849; @RN4887]. Student list products offer many geographic filters. Some geographic filters are based on known geographic borders (e.g., zip code, county, CBSA, state). College Board also created geographic borders that subsequently became filters in the *Student Search Service*, for example the Geodemographic Segment filters.[^12] This manuscript focuses on Geomarkets. Drawing from @RN4786, we argue that geomarkets satisfy the two criteria of racialized inputs. First, they are ostensibly race-neutral inputs in that neither @RN4982, nor subsequent promotional material [e.g., @RN1621], mention race. Second, Geomarkets may be systematically correlated with race. For example, Geomarket "CA07" is "City of Oakland" and "CA08" is "Alameda Country, excluding Oakland." A university might purchase prospect contact information by filtering on CA08 in combination with additional filters, such as year of high school graduation and SAT test score. This article extends @RN5012 by examining how the racial composition of actual student list purchases changes if universities filter on particular Geomarkets.

@@

@RN4982 [pp. 42-44] discuss how the Market Segment Model and Geomarkets should be utilized in the context of purchasing student lists. The authors ask,

> Imagine for the moment that you recruit for a college that draws most of its students from regional or national segments. Where would you concentrate your energies? Ideally, you would seek communities \[Geomarkets\] with a high proportion of students already predisposed toward institutions such as your own. The Market Segment Model would provide this information through segment percentages for the community in question. Further classification of students by social attributes allows you to identify a group for mailings or recruiting\ldots If you were to recruit in Boston, only about two out of every ten students with fewer than two attributes would likely listen, while slightly less than half of the students with two or more attributes would be receptive\ldots. Your efforts would surely be better directed toward \ldots Manchester, Hartford, and Fairfield County \ldots Indeed, in Fairfield County alone you could reach more than 40 percent of your "primary target" population -- that is, students with a greater than 75-percent probability of concentrating their college choices among institutions like your own.

In both mailings and off-campus recruiting visits, @RN4982 recommend focusing on Geomarkets with large populations of students who are deciding between "institutions like your own" (p. 44). Selective colleges primarily draw from students in the "regional" and "national" segments. @RN4982 suggest that the student list purchases of selective colleges should focus on Geomarkets with large numbers of affluent, highly educated households. By contrast, the student list purchases of local state colleges and community colleges, which rely on enrollment from students in the "local" and "in-state," should focus on nearby Geomarkets with large numbers of low-SES students. Our analyses attempt to recreate the strategy recommended by @RN4982 by showing who is included/excluded when student list purchases filter on particular Geomarkets. Considering the correlation between race and wealth in segregated America, we suggest that a class-based list-buying strategy will increase racial stratification in college access. Furthermore, we anticipate that the students of color from affluent Geomarkets will tend to have college-educated parents and be enrolled in affluent, predominantly white schools.
